Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2015-01-20 paper/sheaf_lics.tex
complete pass

As a prerequisite, we encourage the reader to be familiar with type theory and
in particular the point of view developed
in~\cite{hottbook}. Nevertheless, we recall most of the central definitions 
that we use so that the paper is sufficiently self-contained.
%
We note $\prod_{x:T} U$ for dependent products, $\sum_{x:T} U$ for
dependent sums, and $\pi_1, \pi_2$ for the first and second
projection of a dependent pair. We use informal mathematical language
instead of type theory whenever it is possible, to ease the
reading without making our statement imprecise. In particular,
(higher) inductive types are defined using itemization to avoid an
overhead of notation.

@@

As explained in the introduction, a Lawvere-Tierney topology for a
topos $\E$, as defined in Section~\ref{sec:lawv-tiern-sheaf},
corresponds to a left exact modality at level $-1$
%
and sheafification corresponds to lifting this modality to a left
exact modality at level $0$.
%
In homotopy type theory, it is possible to use similar ideas to lift
any left exact modality at level $-1$ to a left exact modality at level
$n$ using sheafification, by induction on $n$.

@@

\begin{defi}
  Let $E$ be a type, and $\chi:E \to \Type_n$. The subobject $\iota :
  A \to E$ of $E$
  classified by $\chi$ is {\em dense} in $E$ when its $\modal_n$-closure
  is equvialent to $\chi_E$, \ie
  %
  $$\forall e:E,~ \left(\sum_{e':E} e=e'\right) \simeq \modal_n
  (\chi~e).$$
  % 
  and moreover, for any $x:A$, the following diagram 
  commutes, 
  $$\xymatrix{
    \sum_{e':A} x = e' \ar@{=}[r] \ar[d]_{\iota'} & (\chi~x) \ar[d]^{\eta_{(\chi~x)}}\\
    \sum_{e':E} x = e' \ar@{=}[r] & \modal_n (\chi~x)
  }$$
  where $\iota': x \mapsto (\iota (\pi_1 x) ; \pi_2 x)$.
\end{defi}

@@

\begin{defi}
  For any type $A$ and $E$, $\iota : A \to E$ and $F:\Type_{n+1}$, we define
    $$
    \Phi_E^\iota : (E \to F) \to (A \to F) 
    $$
    %
    as the map sending an
    arrow $f:E\to F$ to its restriction the restriction of $f \circ m$.
  % \item if $A \to E$ with $n$-truncated homotopy fibers, $\Phi_E^\chi$ is the
  %   map sending $f:~E \to f$ to $\restriction f A$.
  % \end{itemize}
\end{defi}

@@

A fundamental property in the construction of the sheafification
functor is the fact that the type of modal types at level $n$ defines
a sheaf.
%
This property initially holds in the topos-theoretic construction, but
it requires much work here, as we have to deal with equalities. In
particular, this property uses the specific definition
of separated types and sheaves, which shows that we had actually no
choice in their statement.

\begin{prop}\label{prop:sheaf-is-sheaf}
  $\Type_n^\modal$ is a sheaf.
\end{prop}
\begin{proof}
  First, we prove that $\Type_n^\modal$ is separated. Let $E:\Type$
  and $\chi:E \to \Type_n$, dense in $E$. Let $\phi_1,\phi_2:E \to
  \Type_n^\modal$, such that $\phi_1 \circ \pi_1 = \phi_2 \circ
  \pi_1$ and let $x:E$. We show $\phi_1 x = \phi_2 x$ using
  univalence.

@@

Note that once the first modality $\modal_{-1}$ is defined, its
extension to all $\Type_n$ is automatic. This means that the new
logical principles that can be added by the sheaf construction must be
added in the first modality already. Sheafification is just a way
to propagate those new principles to types than are not mere
propositions.

@@

  Let $k:\nat$, we will show that 
  \begin{align*}
    &\sum_{t:T^k} (\mu_T t_1 = \mu_T t_2 \land \cdots
      \land \mu_T t_{k-1} \mu_T t_k)\\
    &= \sum_{t:T^k} \modal_n (t_1 = t_2 \land \cdots
      \land t_{k-1} = t_k)
  \end{align*}
By induction on $k$, and the preservation of products by $\modal_n$, it
suffices to show that for any $a,b:T$, $\modal_n (a=b) = (\mu_T a =
\mu_T b)$. By univalence, we want arrows in both ways, forming an
equivalence.
\begin{itemize}
\item Suppose $p : (\mu_T a = \mu_T b)$. Then projecting $p$ along
  first components yields $q : \prod_{t:T} \modal_n(a=t) = \modal_n (b=t)
  $.
  Taking for example $t=b$, we deduce $\modal_n (a=b) = \modal_n(b=b)$,
  and the latter is inhabited by $\eta_{b=b} 1$.
\item Suppose now $p : \modal_n(a=b)$. Let $\iota$ be the first
  projection from $\separated T \to (T \to \Type_n^\modal)$. $\iota$ is
  an embedding, thus it suffices to prove $\iota (\mu_T a) = \iota
  (\mu_T b)$, \ie $\prod_{t:T}\modal_n (a=t) = \modal_n (b=t)$. The latter
  remains true by univalence.
\end{itemize}
The fact that these two form an equivalence is technical, we refer to
the formalization for an explicit proof.
\end{proof}

@@

To show that $\separated$ is a modality, it remains to show point {\it (v)}. Let $A:\Type_{n+1}$ be a sheaf and $B:A \to \Type_{n+1}$ be
a sheaf family. We want to show that $\sum_{x:A} (Bx)$ is
separated. Let $E$ be a type, and $\sum_{e:E} (\chi\,e)$ a dense
subobject of E.

@@

As this modality is just a step in the construction, we do not show
that it is left exact, we will only do it for the sheafification modality.

@@

%\nt{put the definition of $\modal_{n+1}$ upfront}
For any $T$ in $\Type_{n+1}$, 
$\modal_{n+1}T$ is defined as the closure of $\separated T$,
seen as a subobject of $T \to \Type_n^\modal$. 
%
Explicitly
$$
\modal_{n+1} T \ \defeq \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left\| \sum_{a:T} 
            (\lambda t,~\modal_n (a=t)) = u\right\|.
$$

@@

It remains to show that $\modal_{n+1}$ is left exact and is compatible
with $\modal_{-1}$. To do that, we need to extend the notion of
compatibility and that actually every modality $\modal_{n+1}$ is
compatible with $\modal_n$ on lower homotopy types.
\begin{prop}
  If $T:\Type_n$, then $\modal_{n+1} \widehat T = \modal_n T$, where $\widehat T$ is $T$ seen as a
  $\Type_{n+1}$.
\end{prop}
\begin{proof}
  We prove it by induction on $n$:
  \begin{itemize}
  \item For $n=-1$: Let $T:\HProp$. Then
    \begin{align*}
      \modal_{0} \widehat T &= \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left\| \sum_{a:T} 
      (\lambda t,~\modal_{-1} (a=t)) = u\right\| \\
      &= \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left( \sum_{a:T} 
      (\lambda t,~\modal_{-1} (a=t)) = u\right)
    \end{align*}
    because the type inside the truncation is already in $\HProp$.
    Now, let define $\phi : \modal_{-1} T \to \modal_0T$ by
    $$\phi t = (\lambda t',\, \modal_{-1}(t=t') ; \eta
    (t;\mathrm{idpath}))$$
    and $\psi : \modal_0T\to \modal_{-1} T$ by obtaining the
    witness $a:T$ (which is possible because we are trying to inhabit
    a modal proposition), and letting
    $$\psi (u;x) = \eta_T a.$$
    These two maps form an equivalence (the section and retraction are
    trivial because the equivalence is between mere propositions).
  \item Suppose now that $\modal_{n+1}$ is compatible with all $\modal_k$ on
    lower homotopy types. Let $\modal_{n+2}$ be as above, and let
    $T:\Type_{n+1}$. Then
    \begin{align*}
      \modal_{n+2} \widehat  T &= \hspace{-1em} \sum_{u:T \to \Type_{n+1}^\modal}
                       \hspace{-1em} \modal_{-1}\left\| \sum_{a:T} 
            (\lambda t,~\modal_{n+1} (a=t)) = u\right\| \\
                     &= \hspace{-1em} \sum_{u:T \to
                       \Type_{n+1}^\modal} 
\hspace{-1em} \modal_{-1}\left\| \sum_{a:T} 
            (\lambda t,~\modal_{n} (a=t)) = u\right\|
    \end{align*}
    because $\modal_{n+1}$ is compatible with $\modal_{n}$, and
    $(a=t)$ is in $\Type_n$.

    It remains to prove that for every $(u,x)$ inhabiting the
    $\Sigma$-type above, $u$ is in $T\to\Type_n^\modal$, \ie that for
    every $t:T$, $\IsType n (u\, t)$.  But for any truncation index $p$,
    the predicate $\IsType p:\HProp$ is a a sheaf, so we can get rid
    of $\modal_{-1}$ and of the truncation, which tells us that for
    every 
    $t:T$, $u\, t = \modal_n(a=t) : \Type_n$.
  \end{itemize}
\end{proof}

@@

The last step is the left-exactness of $\modal_{n+1}$. Let
$T$ be in $\Type_{n+1}$ such that $\Contr(\modal_{n+1} T)$.
Thanks to compatibility between $\modal_{n+1}$ and $\modal_n$ for
$\Type_n$, left exactness means that for any $x,y : T$, $\modal_n(x=y)$ is contractible.

@@

Starting from any left-exact modality $\modal_{-1}$ on $\HProp$, we
have defined for any truncation level $n$, a new left-exact modality
$\modal_n$ on $\Type_n$, which is exactly $\modal_{-1}$ when
restricted to $\HProp$.

@@

\begin{prop}
  $(\modal_{\lnot\lnot})_n$, the modality obtained by sheafification
  of the double negation modality,
  %
  induces a type theory where the propositional excluded middle holds.
\end{prop}

@@

A Coq formalization is available at
\url{https://github.com/KevinQuirin/sheafification}.
%
It is based on the Coq/HoTT library available at
\url{https://github.com/HoTT/HoTT} and uses the version of Coq
designed for that library.
%

\subsection{Content of the formalization}

@@

\subsection{Limitations of the formalization}

In our formalization of sheafification defined in
Section~\ref{sec:sheafification} (done in 
\texttt{sheaf\_induction.v}), two very technical facts required to
show that $\separated$ is a modality have been
admitted. 
%
It is due to a lack of technical lemmas on paths between paths in
the Coq/HoTT library.
%
Nevertheless, we are confident that with an enrichment of the Coq/HoTT
library, they can be proved and have been admitted only by lack of
time.
%
In the proof that diagrams are equal in lemma~\ref{lem:sepiscolim},
the equality of projections is admitted, because if this proof is
obvious on paper, it becomes really technical when formalized.
%
This suggest that we should try to fine a better way to encode
projections, with a better computational content.
%
As a consequence, the fact that $P$ is a colimit of the diagonal
closed diagram is admitted too, but again, this proof should be really
short as it is on paper.

More importantly, there are few axioms in the formalization that are
due to an unsatisfactory management of polymorphic universe level in
Coq.
%
Universes are used in type theory to ensure consistency by checking
that definitions are well-stratified according to a certain hierarchy.
%
Universe polymorphism~\cite{sozeau2014universe} supports generic
definitions over universes, reusable at different levels.
%
Although the presence of universe polymorphism is mandatory for our
formalization, its implementation is still to rigid to allow a
complete formalization of our work.

%
If Coq handles cumulativity on all $\Type$ natively, it is not
the case for the $\Sigma$-type $\Type_n$, which require propositional
resizing. 
%
This issue could be solved by adding an axiom of cumulativity
for $\Type_n$ with an explicit management of universes. 
%
But as it would not have any computational content, such a solution
would really complicate the proofs as the axiom would appear
everywhere cumulativity is needed and it would need explicitly
annotation for universe levels everywhere in the formalization.
%
We have chosen a more simple but maybe less satisfactory solution. 
%
When the absence of cumulativity on $\Type_n$ rises a universe issue
for a lemma, we have proved the lemma and then admit the exact same
lemma, just letting Coq infer less constrained universe levels.

An other issue with universe polymorphism lies in the management of
recursive definitions. Indeed, the following recursive definition of
sheafification
%
\[ \begin{array}{l}
   \modal : \forall \ (n : nat), \ \Type_n \to \Type_n 
   \\
    \modal_{-1\phantom{n}} \defeq \lambda T:\HProp,~ \lnot\lnot T \\
      \displaystyle{\modal_{n+1} \defeq  \lambda T:\Type_{n+1}},\\
      \hspace{5em} \displaystyle{\sum_{u:T \to \Type_n^\modal} \modal_{-1} 
      \left\|
      \sum_{a:T} u= (\lambda t,~\modal_n (a=t))
      \right\|}
    \end{array}
\]
%
is not allowed. 
%
This is because Coq forces the universe of the first $\Type_n$
occurring in the definition to be the same for every $n$, whereas the
universe of the first $\Type_{n+1}$ occurring in $\modal_{n+1}$ should be at
least one level higher as the one of $\Type_n$ occurring in
$\modal_{n}$ because of the use of $\Sigma$-type over
$T \to \Type_n^\modal$.
%
% \nt{explain why it is an issue.}
%
Thus, the induction step presented in this paper has been formalized,
but the complete recursive sheafification can not be defined for the
moment.
%
This restriction in our formalization can not be solved without
generalizing universe polymorphism for recursive definition.
%

@@

In this paper, we have defined a way to extend homotopy type theory
with new principles, leveraging Lawvere-Tierney sheafification to
higher topos theory. 
%
Using sheafification, it is possible to add new logical principles to
homotopy type theory, like the law of excluded middle, and this
without axiom.
%
Beyond this result, our work is part of a more general line of works
which aim to illustrate that homotopy type theory is a very promising
theory for formalizing mathematics inside a proof assistant.

@@

We would like to thank all maintainers of the Coq/HoTT library, who
provide a very solid base for our formalization. We also would like
to thank Bas Spitters for helpful discussions on the subject.

>>>>

As a prerequisite, we encourage the reader to be familiar with type theory and
in particular the point of view developed
in~\cite{hottbook}. Nevertheless, we recall most of the central definitions 
that we use so that the paper is sufficiently self-contained.
%
Given a type $T$ and a type family $U : T \to \Type$, we note
$\prod_{x:T} U x$ for the dependent product, $\sum_{x:T} U x$ for the dependent
sum, and $\pi_1, \pi_2$ for the first and second projection of a
dependent pair (noted $(a;b)$). We use informal mathematical language
instead of type theory whenever it is possible, to ease the reading
without making our statement imprecise. In particular, (higher)
inductive types are defined using itemization to avoid an overhead of
notation.

@@

As explained in in Section~\ref{sec:lawv-tiern-sheaf}, a
Lawvere-Tierney topology for a topos $\E$ corresponds to a left exact
modality at level $-1$
%
and sheafification corresponds to lifting this modality to a left
exact modality at level $0$.
%
In homotopy type theory, it is possible to use similar ideas to lift
any left exact modality at level $-1$ to a left exact modality at level
$n$ using sheafification, by induction on $n$.

@@

\begin{defi}
  Let $E$ be a type, and $\chi:E \to \Type_n$. The subobject $\iota :
  A \to E$ of $E$
  classified by $\chi$ is {\em dense} in $E$ when its $\modal_n$-closure
  is equvialent to $\chi_E$, \ie
  %
  $$\forall e:E,~ \left(\sum_{e':E} e=e'\right) \simeq \modal_n
  (\chi~e).$$
  % 
  and moreover, for any $x:A$, the following diagram 
  commutes, 
  $$\xymatrix{
    \sum_{e':A} x = e' \ar@{=}[r] \ar[d]_{\iota'} & (\chi \ x)
    \ar[d]^{\eta_{(\chi \ x)}}\\
    \sum_{e':E} x = e' \ar@{=}[r] & \modal_n (\chi \ x)
  }$$
  where $\iota': x \mapsto (\iota (\pi_1 x) ; \pi_2 x)$.
\end{defi}

@@

\begin{defi}
  For any type $A$ and $E$, $\iota : A \to E$ and $F:\Type_{n+1}$, we define
    $$
    \Phi_E^\iota : (E \to F) \to (A \to F) 
    $$
    %
    as the map sending an
    arrow $f:E\to F$ to its restriction $f \circ m$.
  % \item if $A \to E$ with $n$-truncated homotopy fibers, $\Phi_E^\chi$ is the
  %   map sending $f:~E \to f$ to $\restriction f A$.
  % \end{itemize}
\end{defi}

@@

A fundamental property in the construction of the sheafification
functor is the fact that the type of modal types at level $n$ defines
a sheaf.
%
This property initially holds in the topos-theoretic construction, but
it requires much work here, as we have to deal with equalities. In
particular, this property uses the specific definitions
of separated types and sheaves, which has constituted a guide in the
choice of their precise statements.

\begin{prop}\label{prop:sheaf-is-sheaf}
  $\Type_n^\modal$ is a sheaf.\footnote{Note that we omit the $n$
on $\modal_n$ in $\Type_n^\modal$ to ease the reading.}
\end{prop}
\begin{proof}
  First, we prove that $\Type_n^\modal$ is separated. Let $E:\Type$
  and $\chi:E \to \Type_n$, dense in $E$. Let $\phi_1,\phi_2:E \to
  \Type_n^\modal$, such that $\phi_1 \circ \pi_1 = \phi_2 \circ
  \pi_1$ and let $x:E$. We show $\phi_1 x = \phi_2 x$ using
  univalence.

@@

Note that once the first modality $\modal_{-1}$ is defined, its
extension to all $\Type_n$ is automatic. This means that the new
logical principles that can be added by the sheaf construction must be
added in the first modality already. Sheafification is just a way
to propagate those new principles to types that are not mere
propositions.

@@

  Let $k:\nat$, we have to show that 
  \begin{align*}
    &\sum_{t:T^k} (\mu_T t_1 = \mu_T t_2 \land \cdots
      \land \mu_T t_{k-1} \mu_T t_k)\\
    &= \sum_{t:T^k} \modal_n (t_1 = t_2 \land \cdots
      \land t_{k-1} = t_k)
  \end{align*}
By induction on $k$, and the preservation of products by $\modal_n$, it
suffices to show that for any $a,b:T$, $\modal_n (a=b) = (\mu_T a =
\mu_T b)$. By univalence, we want arrows in both ways, forming an
equivalence.
\begin{itemize}
\item Suppose $p : (\mu_T a = \mu_T b)$. Then projecting $p$ along
  first components yields $q : \prod_{t:T} \modal_n(a=t) = \modal_n (b=t)
  $.
  Taking for example $t=b$, we deduce $\modal_n (a=b) = \modal_n(b=b)$,
  and the latter is inhabited by $\eta_{b=b} 1$.
\item Suppose now $p : \modal_n(a=b)$. Let $\iota$ be the first
  projection from $\separated T \to (T \to \Type_n^\modal)$. $\iota$ is
  an embedding, thus it suffices to prove $\iota (\mu_T a) = \iota
  (\mu_T b)$, \ie $\prod_{t:T}\modal_n (a=t) = \modal_n (b=t)$. The latter
  remains true by univalence.
\end{itemize}
The fact that these two form an equivalence is technical, we refer to
the formalization for an explicit proof.
\end{proof}

@@

To show that $\separated$ is a modality, it remains to show point
{\it (v)} of the definition.
%
Let $A:\Type_{n+1}$ be a sheaf and $B:A \to \Type_{n+1}$ be a sheaf
family. We want to show that $\sum_{x:A} (Bx)$ is separated. Let $E$
be a type, and $\sum_{e:E} (\chi\,e)$ a dense subobject of E.

@@

As this modality is just a step in the construction, we do not need to
show that it is left exact, we will only do it for the sheafification
modality.

@@

%\nt{put the definition of $\modal_{n+1}$ upfront}
For any $T$ in $\Type_{n+1}$, 
$\modal_{n+1}T$ is defined as the closure of $\separated T$,
seen as a subobject of $T \to \Type_n^\modal$. 
%
$\modal_{n+1}T$ can be given explicitly by
$$
\modal_{n+1} T \ \defeq \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left\| \sum_{a:T} 
            (\lambda t,~\modal_n (a=t)) = u\right\|.
$$

@@

It remains to show that $\modal_{n+1}$ is left exact and is compatible
with $\modal_{-1}$. To do that, we need to extend the notion of
compatibility and show that actually every modality $\modal_{n+1}$ is
compatible with $\modal_n$ on lower homotopy types.
\begin{prop}
  If $T:\Type_n$, then $\modal_{n+1} \widehat T = \modal_n T$, where $\widehat T$ is $T$ seen as a
  $\Type_{n+1}$.
\end{prop}
\begin{proof}
  We prove it by induction on $n$:
  \begin{itemize}
  \item For $n=-1$: Let $T:\HProp$. Then
    \begin{align*}
      \modal_{0} \widehat T &= \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left\| \sum_{a:T} 
      (\lambda t,~\modal_{-1} (a=t)) = u\right\| \\
      &= \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left( \sum_{a:T} 
      (\lambda t,~\modal_{-1} (a=t)) = u\right)
    \end{align*}
    because the type inside the truncation is already in $\HProp$.
    Now, let define $\phi : \modal_{-1} T \to \modal_0T$ by
    $$\phi t = (\lambda t',\, \modal_{-1}(t=t') ; \eta
    (t;\mathrm{idpath}))$$
    and $\psi : \modal_0T\to \modal_{-1} T$ by obtaining the
    witness $a:T$ (which is possible because we are trying to inhabit
    a modal proposition), and letting
    $$\psi (u;x) = \eta_T a.$$
    These two maps form an equivalence (the section and retraction are
    trivial because the equivalence is between mere propositions).
  \item Suppose now that $\modal_{n+1}$ is compatible with all $\modal_k$ on
    lower homotopy types. Let $\modal_{n+2}$ be as above, and let
    $T:\Type_{n+1}$. Then
    \begin{align*}
      \modal_{n+2} \widehat  T &= \hspace{-1em} \sum_{u:T \to \Type_{n+1}^\modal}
                       \hspace{-1em} \modal_{-1}\left\| \sum_{a:T} 
            (\lambda t,~\modal_{n+1} (a=t)) = u\right\| \\
                     &= \hspace{-1em} \sum_{u:T \to
                       \Type_{n+1}^\modal} 
\hspace{-1em} \modal_{-1}\left\| \sum_{a:T} 
            (\lambda t,~\modal_{n} (a=t)) = u\right\|
    \end{align*}
    because $\modal_{n+1}$ is compatible with $\modal_{n}$, and
    $(a=t)$ is in $\Type_n$.

    It remains to prove that for every $(u,x)$ inhabiting the
    $\Sigma$-type above, $u$ is in $T\to\Type_n^\modal$, \ie that for
    every $t:T$, $\IsType n (u\, t)$.  But for any truncation index $p$,
    the predicate $\IsType p:\HProp$ is a sheaf, so we can get rid
    of $\modal_{-1}$ and of the truncation, which tells us that for
    every 
    $t:T$, $u\, t = \modal_n(a=t) : \Type_n$.
  \end{itemize}
\end{proof}

@@

The last step is the left-exactness of $\modal_{n+1}$. Let $T$ be in
$\Type_{n+1}$ such that $\Contr(\modal_{n+1} T)$.  Thanks to the just
shown compatibility between $\modal_{n+1}$ and $\modal_n$ for
$\Type_n$, left exactness means that for any $x,y : T$,
$\modal_n(x=y)$ is contractible.

@@

Starting from any left-exact modality $\modal_{-1}$ on $\HProp$, we
have defined for any truncation level $n$, a new left-exact modality
$\modal_n$ on $\Type_n$, which corresponds to $\modal_{-1}$ when
restricted to $\HProp$.

@@

\begin{prop}
  $(\modal_{\lnot\lnot})_n$, the modality obtained by sheafification
  of the double negation modality,
  %
  induces a type theory where the propositional excluded middle law holds.
\end{prop}

Combined with forcing in type theory~\cite{jaber2012extending}, it
should be possible to lift the proof of independence of the continuum
hypothesis to a classical setting, which is where the continuum hypothesis is
really meaningful.  However, we haven't work out the details and left
this for future work.

@@

A Coq formalization is available at
\url{https://github.com/KevinQuirin/sheafification}.
%
It is based on the Coq/HoTT library available at
\url{https://github.com/HoTT/HoTT} and uses the version of Coq
tailored for that library.
%
After reviewing the content and some statistics about the
formalization in Section~\ref{sec:cont-form}, we present the
limitations of our formalization in Section~\ref{sec:limit-form}, in
particular the issues relative to universe polymorphism.

\subsection{Content of the formalization}
\label{sec:cont-form}

@@

\subsection{Limitations of the formalization}
\label{sec:limit-form}

In our formalization of sheafification defined in
Section~\ref{sec:sheafification} (done in 
\texttt{sheaf\_induction.v}), two very technical facts required to
show that $\separated$ is a modality have been
admitted. 
%
It is due to a lack of technical lemmas on paths between paths in
the Coq/HoTT library.
%
This is not surprising because those lemmas are defined in a
call-by-need approach and there is no notion of a complete set of
lemmas on paths for the moment. 
%
This is the subject of ongoing work, in particular those trying to
define a syntax based on the cubical set model~\cite{bezem2013model}
of Coquand et al.  
%
Nevertheless, we are confident that with an enrichment of the Coq/HoTT
library, they can be proved and have been admitted only by lack of
time.
%
In the proof that diagrams are equal in lemma~\ref{lem:sepiscolim},
the equality of projections is admitted, because if this proof is
obvious on paper, it becomes really technical when formalized.
%
This suggest that we should try to fine a better way to encode
projections, with a better computational content.
%
As a consequence, the fact that $P$ is a colimit of the diagonal
closed diagram is admitted too, but again, this proof should be really
as short as it is on paper.

More importantly, there are few axioms in the formalization that are
due to an unsatisfactory management of polymorphic universe levels in
Coq.
%
Universes are used in type theory to ensure consistency by checking
that definitions are well-stratified according to a certain hierarchy.
%
Universe polymorphism~\cite{sozeau2014universe} supports generic
definitions over universes, reusable at different levels.
%
Although the presence of universe polymorphism is mandatory for our
formalization, its implementation is still to rigid to allow a
complete formalization of our work for the following reasons.

%
If Coq handles cumulativity on $\Type$ natively, it is not
the case for the $\Sigma$-type $\Type_n$, which require propositional
resizing. 
%
This issue could be solved by adding an axiom of cumulativity
for $\Type_n$ with an explicit management of universes. 
%
But as it would not have any computational content, such a solution
would really complicate the proofs as the axiom would appear
everywhere cumulativity is needed and it would need explicitly
annotation for universe levels everywhere in the formalization.
%
We have chosen a more simple but maybe less satisfactory solution. 
%
When the absence of cumulativity on $\Type_n$ rises a universe issue
for a lemma (it occurs twice in the formalization), we have proved the
lemma and then admit the exact same lemma, just letting Coq infer less
constrained universe levels.

An other issue with universe polymorphism lies in the management of
recursive definitions. Indeed, the following recursive definition of
sheafification
%
\[ \begin{array}{l}
   \modal : \forall \ (n : nat), \ \Type_n \to \Type_n 
   \\
    \modal_{-1\phantom{n}} \defeq \lambda T:\HProp,~ \lnot\lnot T \\
      \displaystyle{\modal_{n+1} \defeq  \lambda T:\Type_{n+1}},\\
      \hspace{5em} \displaystyle{\sum_{u:T \to \Type_n^\modal} \modal_{-1} 
      \left\|
      \sum_{a:T} u= (\lambda t,~\modal_n (a=t))
      \right\|}
    \end{array}
\]
%
is not allowed. 
%
This is because Coq forces the universe of the first $\Type_n$
occurring in the definition to be the same for every $n$, whereas the
universe of the first $\Type_{n+1}$ occurring in $\modal_{n+1}$ should be at
least one level higher as the one of $\Type_n$ occurring in
$\modal_{n}$ because of the use of $\Sigma$-type over
$T \to \Type_n^\modal$.
%
% \nt{explain why it is an issue.}
%
Thus, the induction step presented in this paper has been formalized,
but the complete recursive sheafification can not be defined for the
moment.
%
This restriction in our formalization may be solved by
generalizing universe polymorphism for recursive definition
%
or with the use of general resizing rules.

@@

In this paper, we have defined a way to extend homotopy type theory
with new principles, leveraging Lawvere-Tierney sheafification to
higher topos theory. 
%
Using sheafification, it is possible to add new logical principles to
homotopy type theory, such as the law of excluded middle, and this
without axiom.
%
Beyond this result, our work is part of a more general line of work
which aim to illustrate that homotopy type theory is a very promising
theory for formalizing mathematics inside a proof assistant.

@@

We would like to thank all maintainers of the Coq/HoTT library, who
have provided a very solid base for our formalization. We also would like
to thank Bas Spitters for helpful discussions on the subject.
